---
title: "maternal"
author: "Joann"
date: "2025-06-05"
output: html_document
---

```{r}
library(tidyverse)
maternal <- read_csv("https://raw.githubusercontent.com/36-SURE/2025/main/data/maternal.csv")

```

```{r}

View(maternal)

```

```{r}

names(maternal)


str(maternal)

```
```{r}

# Data cleaning

as.integer(maternal$PriorBirthsNowDeceased)

```

```{r}

library(tidyverse)
library(dplyr)

PBND <- sum(maternal$PriorBirthsNowDeceased == 'Unknown')
TU <- sum(maternal$TobaccoUse == 'Unknown')
PPD <- sum(maternal$PrePregnancyDiabetes == 'Unknown')
PPH <- sum(maternal$PrePregnancyHypertension == 'Unknown')

missing <- c(PBND, TU, PPD, PPH)
missing

select(.maternal, 
       rows)

for i %in% maternal:
  for k %in% m:
    
dim(maternal)

```

```{r}

missing_data <- maternal %>%
  filter(PriorBirthsNowDeceased == "Unknown" | 
           TobaccoUse == "Unknown" | 
           PrePregnancyDiabetes == "Unknown" | 
           PrePregnancyHypertension == "Unknown") |>
  select(State, PriorBirthsNowDeceased, TobaccoUse, PrePregnancyDiabetes, PrePregnancyHypertension)

Yes_data <- maternal %>%
  filter(PriorBirthsNowDeceased == "Yes" | 
           TobaccoUse == "Yes" | 
           PrePregnancyDiabetes == "Yes" | 
           PrePregnancyHypertension == "Yes") |>
  select(State, PriorBirthsNowDeceased, TobaccoUse, PrePregnancyDiabetes, PrePregnancyHypertension)

No_data <- maternal %>%
  filter(PriorBirthsNowDeceased == "No" | 
           TobaccoUse == "No" | 
           PrePregnancyDiabetes == "No" | 
           PrePregnancyHypertension == "No") |>
  select(State, PriorBirthsNowDeceased, TobaccoUse, PrePregnancyDiabetes, PrePregnancyHypertension)


missing_data

missing_data |>
  ggplot(aes(x=State)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90)) # Rotate 90 degrees for vertical labels

Yes_data |>
  ggplot(aes(x=State)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90))+
  arrange(desc())

No_data |>
  ggplot(aes(x=State)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90))
```


```{r}

tobacco_use <- maternal %>%
  select(State, TobaccoUse) |>
  group_by(State, TobaccoUse) |>
  summarize(count=n(), .groups = "drops")

tobacco_use

tobacco_use <- tobacco_use %>%
  group_by(State) %>%
  mutate(total = sum(count)) %>%
  ungroup()

tobacco_use <- tobaccou_use |>
  mutate(State = reorder(State, -total))


  ggplot(tobacco_use, aes(x= State, y = count, fill = TobaccoUse)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(title = "Counts of Unique Values by Location",
       x = "Location",
       y = "Count") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  


#theme(axis.text.x = element_text(angle = 90)) # Rotate 90 degrees for vertical labels

```


```{r}

missing_data |>
  group_by(State) |>
  summarize(PriorBirthsNowDeceased == "Unknown" | 
           TobaccoUse == "Unknown" | 
           PrePregnancyDiabetes == "Unknown" | 
           PrePregnancyHypertension == "Unknown") 

```

```{r}
  maternal |>
    select(State, 
           PriorBirthsNowDeceased, 
           TobaccoUse, 
           PrePregnancyDiabetes, 
           PrePregnancyHypertension)|>
    filter(PriorBirthsNowDeceased == "Unknown" | 
           TobaccoUse == "Unknown" | 
           PrePregnancyDiabetes == "Unknown" | 
           PrePregnancyHypertension == "Unknown") |>
    rowSums()

```

```{r}
 maternal |>
  select(PriorBirthsNowDeceased, TobaccoUse, PrePregnancyDiabetes, PrePregnancyHypertension) |>
  ggplot(aes(y = 'Yes')) +
   stat_count(
     mapping = NULL,
     geom = "bar",
     show.legend = NA
     )
 
 maternal |>
  select(State, TobaccoUse) |>
   mutate(YesTob == TobaccoUse["Yes"]) |>
   ggplot(aes(x=State, y = YesTob))+
   geom_line()+
   facet_wrap(~stat, ncol=1)


# maternal |>
  # fill(PriorBirthsNowDeceased, .direction = "downup")

```

```{r}

# 
maternal |>
  ggplot(aes(x=AverageNumberofPrenatalVisits)) +
  geom_bar()

```



```{r}

# install.packages('leaflet')
library(leaflet)
# install.packages('maps')
library(maps)
library(ggplot2)
library(dplyr)
# install.packages("mapproj")
library(mapproj)

```

```{r}


num_data <- maternal %>% 
  select(State, Births, 
         AverageMotherAge, 
         AverageBirthWeight, 
         AveragePrePregnancyBMI, 
         AverageNumberPrenatalVisits)

num_data |>
  ggplot(aes(x=AverageBirthWeight))+
  geom_histogram()


(num_data)
```

```{r}



num_data |>
  ggplot(aes(x = State)) +
  geom_histogram()


```

```{r}

library(dplyr)

# Encode categorical variables: Yes = 1, No = 0, Unknown = NA (or 2 if you prefer)
maternal_clean <- maternal %>%
  mutate(
    TobaccoUse_bin = case_when(
      TobaccoUse == "Yes" ~ 1,
      TobaccoUse == "No" ~ 0,
      TobaccoUse == "Unknown" ~ NA_real_
    ),
    Diabetes_bin = case_when(
      PrePregnancyDiabetes == "Yes" ~ 1,
      PrePregnancyDiabetes == "No" ~ 0,
      PrePregnancyDiabetes == "Unknown" ~ NA_real_
    ),
    Hypertension_bin = case_when(
      PrePregnancyHypertension == "Yes" ~ 1,
      PrePregnancyHypertension == "No" ~ 0,
      PrePregnancyHypertension == "Unknown" ~ NA_real_
    )
  )


```

```{r}

# Remove rows with NA in binary columns (optional)
maternal_cluster <- maternal_clean %>%
  filter(!is.na(TobaccoUse_bin) & !is.na(Diabetes_bin) & !is.na(Hypertension_bin))


library(FactoMineR)
library(factoextra)

# Select features for PCA
pca_data <- maternal_cluster %>%
  select(TobaccoUse_bin, Diabetes_bin, Hypertension_bin,
         AverageMotherAge, AverageBirthWeight,
         AveragePrePregnancyBMI, AverageNumberPrenatalVisits,
         AverageIntervalSinceLastBirth)

# Run PCA
pca_result <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Visualize PCA
fviz_pca_biplot(pca_result, repel = TRUE, label="none")

```

```{r}

# Decide optimal number of clusters
fviz_nbclust(pca_data, kmeans, method = "wss")

# Run K-means with chosen number of clusters (say, 3)
set.seed(123)
kmeans_result <- kmeans(scale(pca_data), centers = 3, nstart = 25)

# Add cluster label to data
maternal_cluster$cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters on PCA plot
fviz_cluster(kmeans_result, data = scale(pca_data))


```

Tobacco affect pre-pregnancy? (diabetes & hypertension)
```{r}

# Crosstabs
table(maternal_cluster$TobaccoUse_bin, maternal_cluster$Diabetes_bin)
table(maternal_cluster$TobaccoUse_bin, maternal_cluster$Hypertension_bin)

# Chi-squared tests
chisq.test(maternal_cluster$TobaccoUse_bin, maternal_cluster$Diabetes_bin)
chisq.test(maternal_cluster$TobaccoUse_bin, maternal_cluster$Hypertension_bin)


```



# does using tobaccor or having pre-pregnancy diabetes or hypertension lead to greater amount of prenatal visits?

```{r}

library(ggplot2)

# tobacco vs prenatal
ggplot(maternal_cluster, aes(x = factor(TobaccoUse_bin), y = AverageNumberPrenatalVisits)) +
  geom_boxplot() +
  labs(x = "Tobacco Use (0=No, 1=Yes)", y = "Avg. Prenatal Visits")

# diabetes vs prenatal
ggplot(maternal_cluster, aes(x = factor(Diabetes_bin), y = AverageNumberPrenatalVisits)) +
  geom_boxplot() +
  labs(x = "Diabetes (0=No, 1=Yes)", y = "Avg. Prenatal Visits")

# hypertension vs prenatal
ggplot(maternal_cluster, aes(x = factor(Hypertension_bin), y = AverageNumberPrenatalVisits)) +
  geom_boxplot() +
  labs(x = "Hypertension (0=No, 1=Yes)", y = "Avg. Prenatal Visits")


```
- tobacco : not using tobacco has greater prenatal visits
- diabetes - having diabetes has greater amount of prenatal visits
- hypertension - having hypertension leads to greater amount of visits

```{r}

# correlation 
cor.test(maternal_cluster$AverageNumberPrenatalVisits, maternal_cluster$TobaccoUse_bin)
cor.test(maternal_cluster$AverageNumberPrenatalVisits, maternal_cluster$Diabetes_bin)
cor.test(maternal_cluster$AverageNumberPrenatalVisits, maternal_cluster$Hypertension_bin)

```
- tobacco & prenatal : correlation = -0.3343532, p-value = <2.2E-16 --> there is correlation
- diabetes & prenatal : correlation = 0.4447666, p-value = <2.2E-16 --> there is correlation
- hypertension and prenatal: correlation = 0.283653, p-value = 1.44E-14 --> there is correlation
- none of the confidence intervals contain 0 --> there is a relationship

# mother age affect bmi?
```{r}

ggplot(maternal_cluster, aes(x = AverageMotherAge, y = AveragePrePregnancyBMI)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Mother's Age", y = "Pre-Pregnancy BMI")

# Correlation
cor.test(maternal_cluster$AverageMotherAge, maternal_cluster$AveragePrePregnancyBMI)


```




```{r}
# install.packages("geojsonio")

library(shiny)
library(leaflet)
library(dplyr)

# load geojson
us_states <- geojsonio::geojson_read("us-states.geojson", what = "sp")


aggregated <- maternal %>%
  group_by(State) %>%
  summarise(AvgBMI = mean(AveragePrePregnancyBMI, na.rm = TRUE))

# Merge with spatial data
us_states@maternal <- left_join(us_states@maternal, aggregated, by = c("name" = "State"))

ui <- fluidPage(
  titlePanel("Pre-pregnancy BMI by State"),
  leafletOutput("map")
)

server <- function(input, output, session) {
  output$map <- renderLeaflet({
    pal <- colorNumeric(palette = "YlOrRd", domain = us_states@maternal$AvgBMI)

    leaflet(us_states) %>%
      addProviderTiles("CartoDB.Positron") %>%
      addPolygons(fillColor = ~pal(AvgBMI),
                  fillOpacity = 0.7,
                  color = "#BDBDC3",
                  weight = 1,
                  popup = ~paste(name, "<br>", "Avg BMI:", round(AvgBMI, 2))) %>%
      addLegend(pal = pal, values = ~AvgBMI, title = "Avg Pre-Pregnancy BMI")
  })
}

shinyApp(ui, server)

```




