---
title: "R Notebook"
output: html_notebook
---
# Maternal healthcare disparities

## Introduction

This project explores a maternal healthcare dataset sourced from the CDC WONDER program. The dataset contains state-level aggregated birth records for 2019, including information on maternal characteristics (e.g., age, BMI), prenatal care (number of prenatal visits), birth outcomes (birth weight), and risk factors (tobacco use, pre-pregnancy diabetes, pre-pregnancy hypertension, prior infant deaths). Each row represents aggregated data for a state, including a set of key risk-indicator variables with summary statistics.

Improving maternal and infant health is a priority for public health agencies, and this dataset allows us to investigate patterns, disparities, and potential risk groupings that might inform targeted interventions.

---

## Objectives

The goals of this analysis are:

1. **Examine prenatal care by risk factors**  
   Investigate whether mothers with certain risk factors (e.g., smoking, diabetes, hypertension) receive different levels of prenatal care.

2. **Identify key associations**  
   Use correlation analysis and visualizations to find the strongest relationships among numerical variables (e.g., age, BMI, birth weight, prenatal visits, interval since last birth).

3. **Cluster maternal risk profiles**  
   Reduce dimensionality using PCA and apply K-means clustering on the principal components to identify meaningful risk profile groups.

4. **Map state-level risk profiles**  
   Aggregate cluster assignments at the state level and visualize the geographic distribution of similar maternal risk profiles across the U.S.

---

## Project Workflow Overview

---

*Proceed below to explore each step with code, visualizations, and interpretation.*  
```{r}
library(tidyverse)
library(ggcorrplot)
library(cowplot)
library(usmap)
theme_set(theme_light())
maternal <- read_csv("https://raw.githubusercontent.com/36-SURE/2025/main/data/maternal.csv")
```
## Data Preparation
```{r message=FALSE, warning=FALSE}
# Change PriorBirthsNowDeceased to numeric
maternal$PriorBirthsNowDeceased <- as.numeric(maternal$PriorBirthsNowDeceased)

# Handle numerical missing values
# colSums(is.na(maternal))
maternal <- maternal |>
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# Check and remove Unknowns
# colSums(maternal == "Unknown", na.rm = TRUE)
maternal <- maternal |>
  filter_all(all_vars(. != "Unknown"))
```
## Exploratory Data Analysis
```{r}
summary(maternal)
```
### Question1:
Does an increased number of prenatal visits correlate with risk factors (tobacco use, diabetes, hypertension)?
```{r}
# Create pivot table that will allow us to create facets
maternal_visit <- maternal |>
  pivot_longer(
    cols = c(TobaccoUse, 
             PrePregnancyDiabetes, 
             PrePregnancyHypertension),
    names_to = "risk_type",
    values_to = "risk_value"
  )
maternal_visit |> 
  ggplot(aes(x = risk_value,
             y = AverageNumberPrenatalVisits,
             fill = risk_value)) +
  geom_bar(stat = "summary", fun = mean) +
  facet_wrap(~ risk_type, nrow = 1) + 
  labs(
    title = "Average Prenatal Visits by Risk Factor",
    x = "Risk Group",
    y = "Average Prenatal Visits"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", 
        panel.spacing.x = unit(0.5, "cm"),
        plot.title = element_text(hjust = 0.5,
                                  size = 14,
                                  face = "bold"))
```
### Question2:
Are there strong correlations among the numeric variables in the data-set? If so, what are the most important ones, and how might they be related?
```{r}
# First let's have a correlation heat-map to tell us variables that correlated
corr_data <- maternal |>
  select(where(is.numeric))

# Compute the correlation matrix  
corr_matrix <- cor(corr_data, use = "complete.obs")
ggcorrplot(corr_matrix,
           method = "circle",
           lab = TRUE, 
           type = "lower", 
           title = "Correlation Heatmap",
           colors = c("red", "white", "blue")) +
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 14,
                                  face = "bold"))
```
```{r}
# Plot 1: Mother's Age vs Pre-Pregnancy BMI
plot1 <- maternal |>
  ggplot(aes(x = AverageMotherAge, 
             y = AveragePrePregnancyBMI)) +
  geom_point(alpha = 0.5, color = "royalblue") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Mother's Age vs Pre-Pregnancy BMI",
    x = "Average Mother Age",
    y = "Average Pre-Pregnancy BMI"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 14,
                                  face = "bold"))

# Plot 2: BMI vs Prenatal Visits
plot2 <- maternal |>
  ggplot(aes(x = AveragePrePregnancyBMI, 
             y = AverageNumberPrenatalVisits)) +
  geom_point(alpha = 0.5, color = "seagreen") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "BMI vs Prenatal Visits",
    x = "Average Pre-Pregnancy BMI",
    y = "Average Number of Prenatal Visits"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 14,
                                  face = "bold"))

plot_grid(plot1, plot2, nrow = 2)
```
## Clustering analysis
Do certain states form similar risk profiles?
```{r message=FALSE, warning=FALSE}
# Principal Component Analysis
# Binary encoding 
maternal <- maternal |>
  mutate(across(c(TobaccoUse, 
                   PrePregnancyDiabetes, 
                   PrePregnancyHypertension), ~ as.numeric(. == "Yes")))

risk_profile_vars <- maternal |>
  select(-State, -Births, -id)

# Apply PCA
maternal_pca <- prcomp(risk_profile_vars, center = TRUE, scale. = TRUE)

# Create a scree plot
maternal_pca |>
  fviz_eig(addlabels = TRUE) +
  geom_hline(
    yintercept = 100 * (1 / ncol(maternal_pca$x)),
    linetype = "dashed",
    color = "darkred"
  )
```
## Apply K-Means clustering
```{r}
# Extract PC1, PC2, PC3 and PC4
maternal_pca_data <- data.frame(maternal_pca$x[,1:4])

# Determine number of clusters
fviz_nbclust(maternal_pca_data,
             FUNcluster = kmeans,
             method = 'wss')

set.seed(123)
my_kmeans <- kmeans(maternal_pca_data, 
                    centers = 4,
                    nstart = 30)

# Visualize the clusters
fviz_pca_ind(maternal_pca, 
             habillage = my_kmeans$cluster,
             label = "none",
             addEllipses = TRUE)
```
## Visualize clusters by State
```{r}
```


